The C programming language is one of the most widely used programming languages in the world. 
It is used in a wide variety of applications, ranging from embedded systems to cryptographic libraries. 
However, C is also notorious for lacking security guarantees.
Security-related issues in programs written in C stem from both the programmer, but also from the compiler.

The problem with security issues introduced by the compiler is especially prevalent in the context of timing attacks on cryptographic algorithms. 
Developers try to mitigate timing attacks by writing constant-time code -- code where the execution time is independent of the input.
However, the compiler may introduce timing leaks by optimizing away the constant-time code.
Developers try to get around the compiler optimizations by obfuscating their code. 
However, the obfuscated code might still be vulnerable to an optimization that introduces a timing vulnerability.
In a recent study \citep{developer-survey-timing-attacks}, it was shown that the vast majority of developers of prominent open-source cryptographic libraries rely on constant-time code practices that in theory result in constant-time code, but may be vulnerable after compilation.

The issue of timing vulnerabilities introduced by compilers is well-known in the community.
Several proposals have been made to mitigate the issue \citep{dudect, fact,what-you-c, verified-constant-time-c-comiler}.
However, the problem persists and to the best of our knowledge, no study has quantified the issue of timing vulnerabilities introduced by compilers.

In this paper, we try to quantify the issue of timing vulnerabilities introduced by the two most popular C compilers, \texttt{gcc} and \texttt{clang} on the x86-64 architecture.
We take a data-based approach, implementing a tool, OptiFuzz, that generates, fuzzes and analyzes random C programs.
For simplicity, the generated programs are limited to only containing non-branching arithmetic, logical and comparison operations.
During the analysis phase, OptiFuzz applies automatic state-of-the-art statistical analysis of the data to detect timing leaks.
We use our results from OptiFuzz to investigate what optimization flags are responsible for introducing timing leaks and quantify the extent of the problem.
Finally, we discuss how the issue of timing vulnerabilities introduced by the compiler can be mitigated by using language-based security techniques.

\subsection{Related Work}
The issue of timing vulnerabilities introduced by the clang C compiler across different versions has been researched \citep{what-you-c}. 
Several studies have investigated potential solutions to the issue, including using constant-time branching instructions \citep{what-you-c}, using black-box testing software \citep{dudect}, using domain-specific languages \citep{fact}, and notably a verified constant-time C compiler has been developed \citep{verified-constant-time-c-comiler}.

\subsection{Contributions}
We provide a quantitative analysis of timing vulnerabilities introduced by \texttt{gcc} and \texttt{clang}, focusing on specific troublesome optimization flags.
We also provide a tool, OptiFuzz, that can be used to generate, fuzz and analyze random C programs for further investigation of the issue.
Finally, we provide a discussion of how the issue of timing vulnerabilities introduced by the compiler can be mitigated based on our findings.

\subsection{Overview}
This paper is organized as follows: 
In Section \ref{sec:preliminaries}, we provide some background knowledge of timing attacks and optimizing compilers.
In Section \ref{sec:optifuzz}, we provide a detailed description of our tool, and the methodology used to acquire our results.
In Section \ref{sec:results}, we present our experimental results.
In Section \ref{sec:evaluation}, we evaluate our tool and methodology, and the quality of our results.
In Section \ref{sec:discussion}, we discuss various possible solutions to the issue based on other research.
Furthermore, we provide some of our raw results in the appendix.