\section*{Appendix Reading Guide}
The following is a guide on how to interpret the results produced by OptiFuzz. 

In the top of a page, a program signature line will be shown:
\begin{center}
    \textbf{Program 1} -- \texttt{Seed 257634744} -- \texttt{gcc O0 O1 O2 O3}\\\small\texttt{Classes: uniform fixed xzero yzero xlty yltx small equal max64}\\[2mm]
\end{center}
The seed used to generate the program is printed, followed by the compiler and compiler flags.
On a new line, all input classes that the fuzzer used are printed.
Refer to \ref{sec:fuzzing} for a description of each input class.

Next the program is shown:
\begin{lstlisting}[style=defstyle,language=C]
...
int program(long long int x, long long int y) { return (((44 ^ ((false >> (-874552120269441001 == x)) | ((53 * 29) >= (-5509653365268467218 ^ x)))) * 24) & (false + x)); }
\end{lstlisting}
The first two lines of the program have been trimmed. Namely, two define statements
which define \texttt{false} to 0 and \texttt{true} to 1.
What follows now is a histogram of the measured times:
\input{appendix/reading-guide-assets/figure-tikz.tex}
The histogram caption at the top indicates which compiler flag was used to compile the program.
The table below the histogram indicates the means for the different input classes.
These means have been drawn as dashed lines over the histogram.

The histogram plot is not conventional, as the saturation indicates agreement across all input classes.
That is, if a bucket is close to transparent, then that might indicate, that it only came from one input class.
If multiple input classes have bins at the same location with the same height, the bin will be very saturated. 
If we examine the histogram above, we see that some input classes got about a total of about 1000 measurements in the [25, 32] interval.
As no mean sticks out - other than the \texttt{equal} class - it might be disregarded as noise.

Next is the assembly, hypothesis rejection flag, and jump indicators, which should be directly below the histogram.
\input{appendix/reading-guide-assets/asm-lst.tex}
Only if the hypothesis was rejected, it will be marked with red.

The jump indicator \texttt{jg [10]} tells us, that a \texttt{jg} instruction was found on line 10.

\newpage