\documentclass[10pt]{article}
\input{prelude}
\usepackage[authoryear]{natbib}

\title{
  Optimizing Away Security in C
}

\groupNumber{10}
\author{Anders B. Clausen \and Johan T. Degn \and Jonathan S. Eilath}

\begin{document}
\maketitle
\thispagestyle{empty}

\todo{Add short guide to read data plots}
\todo{Make figures smaller and better, and remove [H]}
\todo{Remove section from fuzzing and move to analysis}
\todo{Enable appendix}

\section*{Abstract}
Developers strive to write secure cryptographic code that is resistant to various attacks, including timing attacks.
It is a well-established fact that C compilers can introduce timing vulnerabilities into programs, even when the source code is seemingly secure.
However, to our knowledge, the extent of this problem is quantitatively unknown: 
How bad is the problem?
In this paper, we present OptiFuzz, a tool that can be used to test compilers for timing vulnerabilities.
OptiFuzz has been optimized considering the C language specification, the x86-64 architecture, and state-of-the-art statistical analysis.
This has resulted in a simple tool that can reliably indicate the extent of the real-world issue of timing vulnerabilities introduced by compilers.
Using OptiFuzz, we have generated and compiled 600,000 random C programs and shown that between 0.15\% and 12.63\% of these programs are vulnerable to timing attacks when compiled with \texttt{gcc} depending on the configured level of optimization.
Contrarily, we show that \texttt{clang} introduces timing vulnerabilities much more rarely.
This has led us to pinpoint specific optimizations within \texttt{gcc} that are responsible for introducing timing vulnerabilities, as well as specific expression patterns that are vulnerable to resulting in timing vulnerabilities when compiled.
In light of this result, we have discussed how the issue of timing vulnerabilities introduced by compilers can be mitigated by using language-based security techniques from the literature.

\section{Introduction}
\input{chapters/introduction.tex}

\section{Preliminaries}
\input{chapters/preliminaries.tex}

\section{OptiFuzz}
\input{chapters/optifuzz.tex}
\subsection{Code Generation}
\input{chapters/code-generation.tex}
\subsection{Assembly Inspection}
\input{chapters/assembly-inspection.tex}
\subsection{Fuzzing}
\input{chapters/fuzzing.tex}
\subsection{Statistical Analysis}
\input{chapters/statistical-analysis.tex}

\section{Experimental Results}
\input{chapters/results.tex}

\section{Evaluation}
\input{chapters/evaluation.tex}

\section{Solving the Problem}
\input{chapters/discussion.tex}

\section{Conclusion and Future Work}
\input{chapters/conclusion.tex}

\pagebreak

\bibliographystyle{abbrvnat}
\bibliography{litterature}

\pagebreak
\begin{appendices}

\section{Select General Optimizations Results}
\label{appendix:general-optimizations-results}
%\input{appendix/general-optimizations-results/master.tex}

\end{appendices}

\end{document}